Model Finding for (Co)recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action \relax{Contrexemples Utilisables par Isabelle et Coq} (CUIC). }

### 2015-05-03 tex/main.tex
Minor fixes and comments

\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
\newcommand{\ajr}[1]{\rem{#1 --ajr}}

@@

%%% I added (Co) -- hope you like it --JB
\title{Model Finding for (Co)recursive Functions in SMT}

@@

first-order model finders: SEM, MACE/Mace, Paradox, iProver
  either direct search or reduction to SAT

@@

A \emph{signature} $\Sigma$ consists of 
a set $\ssorts{\Sigma}$ of type symbols and
a set $\sfuns{\Sigma}$ of function symbols.
We assume that signatures always include a Boolean type $\Bool$ and constants 
$\ltrue$ and $\lfalse$ of that type (respectively, for true and false), 
an (infix) equality predicate $\teq$ of type $\tau \times \tau$ for each $\tau \in \ssorts{\Sigma}$,
and standard boolean connectives ($\neg$, $\wedge$, $\vee$, $\lite$, etc.).
and a function $\lite$ of type $\Bool \times \tau \times \tau \rightarrow \tau$ for each $\tau \in \ssorts{\Sigma}$.
We will write $t\typ{\tau}$ to denote a term $t$ of type~$\tau$.
We write $\terms( t )$ to denote the set of subterms in $t$.
%We write $\functerms^\Sigma( t )$ to denote the set of $f$-applications in $t$ such that $f \in \sfundefs{\Sigma}$.

@@

%In this paper, we associate quantified formulas with uninterpreted functions (informally, those whose semantics they define).
In this paper, we consider \emph{annotated quantifed formulas} of the form $\forallf{f} \vec x. \varphi$, 
where $f$ is an uninterpreted function from $\sfuns{\Sigma}$, or the distinguished constant $\fnull$.
An annotated quantified formula $\forallf{f} \vec x. \varphi$ is a \emph{function definition (for $f$)} if 
$f$ is a non-$\fnull$ function of type $\tau_1 \times \ldots \times \tau_n \rightarrow \tau$, 
$\vec x$ is a tuple of $n$ variables of type $\tau_1, \ldots, \tau_n$
and $\forallf{f} \vec x. \varphi$ is eqvuivalent to $\forall \vec x. f( \vec x ) \teq t$ for some $t$.
We will write $\forall \vec x. \varphi$ as shorthand for $\forallf{\fnull} \vec x. \varphi$,
and $\exists \vec x. \varphi$ as shorthand for $\neg \forall \vec x. \neg \varphi$.

\begin{definition}
A formula $\varphi$ is \emph{well-defined with respect to functions} $\{ f_1, \ldots, f_n \}$ if it is of the form:
\begin{eqnarray} \label{eq:wdf}
(\forallf{f_1} \vec x_1. \varphi_1) \wedge \ldots \wedge (\forallf{f_n} \vec x_n. \varphi_n) \wedge \psi
\end{eqnarray}
where $f_1, \ldots, f_n$ are not $\fnull$ and distinct, 
and $\psi$ contains no function definitions.
We will commonly call $\psi$ the \emph{conjecture} of $\varphi$.
\end{definition}

@@

Recent techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} 
have focused on finding models for quantified formulas in SMT.
A instatiation-based procedure is shown in~\cite{GeDeM-CAV-09} to be complete for formulas in the \emph{essentially uninterpreted} fragment.
This fragment is limited to universally quantified formulas where all occurrences of variables occur as direct subchildren of uninterpreted functions.
For instance, the formula $\forall x\typ{\sortint}. f( x ) \teq g( x ) + 5$ 
is in the essentially uninterpreted fragment since all occurrences of $x$ occur beneath uninterpreted functions $f$ and $g$.
Other syntactic criteria are identified in this paper that extend this fragment slightly, including cases when variables occur as arguments of arithmetic predicates.
The techniques in~\cite{ReyEtAl-1-RR-13} can be used for finding finite models for quantified formulas over uninterpreted types as well as types having a 
fixed finite interpretation (such as fixed-width bitvectors, datatype enumerations, etc.).
For instance, 
finite model finding techniques can find a model for
the formula $\forall x\typ{\tau} y\typ{\tau}. \neg x \teq y \vee f( x ) \teq f( y )$
where the uninterpreted type $\tau$ is interpreted as a finite set, and $f$ as injective finite map.

@@

Here, we advocate an approach that translates formulas involving function definitions (such as the one mentioned in the previous paragraph)
into equisatisfiable formulas where the techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are applicable.

@@

In the remainder of the section, we fix a signature $\Sigma$.
Say that $\Sigma$-formula $\varphi$ is well-defined with respect to functions $\sfundefs{\Sigma}$, 
and the set of function definitions corresponding to $\sfundefs{\Sigma}$ are well-specified.
%We may translate $\varphi$ into a equisatisfiable formula $\psi$ for which known model-finding procedures~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are applicable.
This section presents a method for constructing an extended signature $\extendsig{ \Sigma }$,
and a $\extendsig{ \Sigma }$-formula $\psi$ such that $\varphi$ and $\psi$ are equisatisfiable.
The intuition behind this translation is use an uninterpreted type to abstract the set of \emph{relevant} tuples for each recursive function $f$,
and restrict the quantification of the function definition for $f$ to a variable of this type.
Informally, the relevant tuples $\vec t$ of a function $f$ are the ones for which the interpretation of $f( \vec t )$ is relevant to the satisfiability of $\varphi$.

@@

We have seen a satisfiability-preserving translation $\conv$ that significantly improves the effectiveness of known SMT techniques
for finding counterexamples for formulas involving recursive function definitions.
For convienience, the translation $\conv$ has been implemented in the latest development version of \cvc~\cite{ReyEtAl-1-RR-13} (version 1.5 pre-release).
Function definitions $\forallf{f} \vec x. \varphi$ can be written in using the $\definefunreccmd$ command, 
proposed in the draft of SMT version 2.5~\cite{}.
For instance, the input from Example~\ref{ex:translation} can be written as:

\begin{verbatim}
(define-fun-rec s ((x Int)) Int (ite (<= x 0) 0 (s (- x 1))))
(assert (> (s x) 100))
(check-sat)
\end{verbatim}

When reading this input, 
\cvc adds the (annotated) quantified formula $\forallf{s} x. s( x ) \teq \lite( x \leq 0, 0, s( x-1 ) )$ to its list of assertions,
which after rewriting becomes $\forallf{s} x. \lite( x \leq 0, s( x ) \teq 0, s( x ) \teq s( x-1 ) )$.
If \cvc is in finite model finding mode for recursive functions is enabled (using the command line parameter ``--fmf-fun"),
then it will replace its list of known assertions based on the conversion $\conv$ before checking for satisfiability.
Note that this option is model-sound only if the functions provided in the input are well-specified.

@@

{%\footnotesize
\def\ackname{Acknowledgment}
\paragraph{%\footnotesize
\ackname.}
We would like to thank Damien Busato-Gaston and Emmanouil Koukoutos for providing the initial set of benchmarks used in the evaluation.
This research is partially supported by the Inria technological development
action \relax{Contrexemples Utilisables par Isabelle et Coq} (CUIC).

}

>>>>

\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
\newcommand{\ajr}[1]{\rem{#1 --ajr}}
\newcommand{\jb}[1]{\rem{#1 --jb}}
\newcommand{\ct}[1]{\rem{#1 --ct}}

@@

%%% I added (Co) -- hope you like it --JB
\title{Model Finding for (Co)recursive Functions in SMT\thanks{
This research is partially supported by the Inria technological development
action \relax{Contrexemples Utilisables par Isabelle et Coq} (CUIC).
}
}

@@

first-order model finders: SEM, MACE/Mace, Paradox, FMDarwin, iProver
  either direct search or reduction to SAT

@@

A \emph{signature} $\Sigma$ consists of 
a set $\ssorts{\Sigma}$ of type symbols and
a set $\sfuns{\Sigma}$ of function symbols.
We assume that signatures always include a Boolean type $\Bool$ and constants 
$\ltrue$ and $\lfalse$ of that type (respectively, for true and false), 
an (infix) equality predicate $\teq$ of type $\tau \times \tau$ for each $\tau \in \ssorts{\Sigma}$,
and standard boolean connectives ($\neg$, $\wedge$, $\vee$, etc.).
and a function $\lite$ of type $\Bool \times \tau \times \tau \rightarrow \tau$ for each $\tau \in \ssorts{\Sigma}$.
We will write $t\typ{\tau}$ to denote a term $t$ of type~$\tau$.
We write $\terms( t )$ to denote the set of subterms in $t$.
%We write $\functerms^\Sigma( t )$ to denote the set of $f$-applications in $t$ such that $f \in \sfundefs{\Sigma}$.

@@

%In this paper, we associate quantified formulas with uninterpreted functions (informally, those whose semantics they define).
In this paper, we consider \emph{annotated quantifed formulas} of the form $\forallf{f} \vec x. \varphi$, 
where $f$ is an uninterpreted function from $\sfuns{\Sigma}$, or the distinguished constant $\fnull$.
An annotated quantified formula $\forallf{f} \vec x. \varphi$ is a \emph{function definition (for $f$)} if 
$f$ is a non-$\fnull$ function of type $\tau_1 \times \ldots \times \tau_n \rightarrow \tau$, 
$\vec x$ is a tuple of $n$ variables of type $\tau_1, \ldots, \tau_n$
and $\forallf{f} \vec x. \varphi$ is equivalent to $\forall \vec x. f( \vec x ) \teq t$ for some $t$.
We will write $\forall \vec x. \varphi$ as shorthand for $\forallf{\fnull} \vec x. \varphi$,
and $\exists \vec x. \varphi$ as shorthand for $\neg \forall \vec x. \neg \varphi$.

\begin{definition}
A formula $\varphi$ is \emph{well-defined with respect to functions} $\{ f_1, \ldots, f_n \}$ if it is of the form:
\begin{eqnarray} \label{eq:wdf}
(\forallf{f_1} \vec x_1. \varphi_1) \wedge \ldots \wedge (\forallf{f_n} \vec x_n. \varphi_n) \wedge \psi
\end{eqnarray}
where $f_1, \ldots, f_n$ are not $\fnull$ and distinct, 
and $\psi$ contains no function definitions.
We will commonly call $\psi$ the \emph{conjecture} of $\varphi$.
\ct{Don't we need additional restrictions on the $\varphi_i$'s here?
(E.g., they are quantifier-free, or perhaps existential or containing only let binders.)}
\end{definition}

@@

Recent techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} 
have focused on finding models for quantified formulas in SMT.
A instantiation-based procedure is shown in~\cite{GeDeM-CAV-09} to be complete for formulas in the \emph{essentially uninterpreted} fragment.
This fragment is limited to universally quantified formulas where all occurrences of variables occur as direct subchildren of uninterpreted functions.
For instance, the formula $\forall x\typ{\sortint}. f( x ) \teq g( x ) + 5$ 
is in the essentially uninterpreted fragment since all occurrences of $x$ occur beneath uninterpreted functions $f$ and $g$.
Other syntactic criteria are identified in this paper that extend this fragment slightly, including cases when variables occur as arguments of arithmetic predicates.
The techniques in~\cite{ReyEtAl-1-RR-13} can be used for finding finite models for quantified formulas over uninterpreted types as well as types having a 
fixed finite interpretation (such as fixed-width bitvectors, datatype enumerations, etc.).
For instance, 
finite model finding techniques can find a model for
the formula $\forall x\typ{\tau} y\typ{\tau}. \neg x \teq y \vee f( x ) \teq f( y )$
where the uninterpreted type $\tau$ is interpreted as a finite set, and $f$ as injective finite map.

@@

Here, we advocate an approach that translates formulas involving function definitions (such as the one mentioned in the previous paragraph)
into equisatisfiable \ct{to be defined} formulas where the techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are applicable.

@@

In the remainder of the section, we fix a signature $\Sigma$.
Say that $\Sigma$-formula $\varphi$ is well-defined with respect to functions $\sfundefs{\Sigma}$, 
and the set of function definitions corresponding to $\sfundefs{\Sigma}$ are well-specified.
%We may translate $\varphi$ into a equisatisfiable formula $\psi$ for which known model-finding procedures~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are applicable.
This section presents a method for constructing an extended signature $\extendsig{ \Sigma }$,
and a $\extendsig{ \Sigma }$-formula $\psi$ such that $\varphi$ and $\psi$ are equisatisfiable.
The intuition behind this translation is to use an uninterpreted type \ct{to be defined} to abstract the set of \emph{relevant} tuples for each recursive function $f$,
and restrict the quantification of the function definition for $f$ to a variable of this type.
Informally, the relevant tuples $\vec t$ of a function $f$ are the ones for which the interpretation of $f( \vec t )$ is relevant to the satisfiability of $\varphi$.

@@

We have seen a satisfiability-preserving translation $\conv$ that significantly improves the effectiveness of known SMT techniques
for finding counterexamples for formulas involving recursive function definitions.
For convienience, the translation $\conv$ has been implemented in the latest development version of \cvc~\cite{ReyEtAl-1-RR-13} (version 1.5 pre-release).
Function definitions $\forallf{f} \vec x. \varphi$ can be written in using the $\definefunreccmd$ command, 
from Cersion 2.5~\cite{} of the SMT-LIB standard.
For instance, the formula $\varphi$ in Example~\ref{ex:translation} 
can be written as:

{\small
\begin{verbatim}
(define-fun-rec s ((x Int)) Int (ite (<= x 0) 0 (s (- x 1))))
(assert (> (s x) 100))
(check-sat)
\end{verbatim}
}

When reading this input, 
\cvc adds the (annotated) quantified formula $\forallf{s} x. s( x ) \teq \lite( x \leq 0, 0, s( x-1 ) )$ to its list of assertions,
which after rewriting becomes $\forallf{s} x. \lite( x \leq 0, s( x ) \teq 0, s( x ) \teq s( x-1 ) )$.
If \cvc's finite model finding mode for recursive functions is enabled (using the command line parameter ``--fmf-fun"),
then it will replace its list of known assertions based on the conversion $\conv$ before checking for satisfiability.
Note that this option is model-sound \ct{define?} only if the functions provided in the input are well-specified.

@@

{%\footnotesize
\def\ackname{Acknowledgment}
\paragraph{%\footnotesize
\ackname.}
We would like to thank Damien Busato-Gaston and Emmanouil Koukoutos for providing the initial set of benchmarks used in the evaluation.
}
